---
title: "TidyTuesday June 16, 2026"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hello!

This is my **first** time working with Tidy Tuesday data. I'm a beginning student in a Master's in Data Science program. Wish me luck!🍀

### Load the necessary packages

```{r loadPackages, message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
library(knitr)
```

### Load the data


```{r loadData}
# Load the data
eng_wales <- read_csv("16Jun2026 data/raw_england_wales_names.csv")
nor_ire <- read_csv("16Jun2026 data/raw_ni_names.csv")
scot <- read_csv("16Jun2026 data/raw_scotland_names.csv")

```

## Data analysis

### What's in the data?

```{r dataLooks}
# Find range of years in each dataset
print("Range for England_Wales:")
range(eng_wales$Year, na.rm = TRUE)
print("Range for North_Ireland")
range(nor_ire$Year, na.rm = TRUE)
print("Range for Scotland")
range(scot$Year, na.rm = TRUE)


```
A quick glance at the data shows they do not share the same time frame. The English_Wales dataset includes years from 1996-2024. The North_Ireland dataset includes years from 1997-2025. The Scotland dataset includes years from 1997-2025. Let's limit the data to a 5-year window between 2020 and 2024.

```{r fiveYears}
eng_wales5 <- filter(eng_wales, Year >= 2020 & Year <= 2024)
nor_ire5 <- filter(nor_ire, Year >= 2020 & Year <= 2024)
scot5 <- filter(scot, Year >= 2020 & Year <= 2024)

# Get the overall summary of each dataset
print("English_Wales Data Summary:")
summary(eng_wales5)
print("North_Ireland Data Summary:")
summary(nor_ire5)
print("Scotland Data Summary:")
summary(scot5)

# Remove NA values in nor_ire5 dataset
nor_ire5 <- nor_ire5 %>% drop_na()

# Verifying NAs were dropped
nrow(nor_ire5)
```

```{r customRanking}

# Get the total number of names for each dataset
total_ew <- sum(eng_wales5$Number)
total_ni <- sum(nor_ire5$Number)
total_s <- sum(scot5$Number)

# Create a new column in each data set that holds a custom-calculated ranking
eng_wales5 <- eng_wales5 %>% mutate(Custom_Rank = (Number / total_ew) * 100)
nor_ire5 <- nor_ire5 %>% mutate(Custom_Rank = (Number / total_ni) * 100)
scot5 <- scot5 %>% mutate(Custom_Rank = (Number / total_s) * 100)

```


### Let's try some plots...


```{r columnEngWales5}
# Plot the top 10 baby names using a column chart
eng_wales5 %>% slice_max(Custom_Rank, n = 10) %>% ggplot(aes(x = Custom_Rank, y = Name)) + geom_col()
```

```{r columnNorWales5}
# Plot the top 10 baby names using a column chart
nor_ire5 %>% slice_max(Custom_Rank, n = 10) %>% ggplot(aes(x = Custom_Rank, y = Name)) + geom_col()
```


```{r columnScot5}
# Plot the top 10 baby names using a column chart
scot5 %>% slice_max(Custom_Rank, n = 10) %>% ggplot(aes(x = Custom_Rank, y = Name)) + geom_col()
```

It's **interesting** that the only dataset that has girl names in the top 10 ranking is Scotland. Is it possible that there were more girls born in Scotland than the other two regions?

### Check 👦 vs 👧 babies born in the 5-year time frame for each region

```{r Boy-v-Girl}
# Find England_Wales total sex counts
ew_boy_total <- sum(eng_wales5$Number[eng_wales5$Sex == "Boy"], na.rm = TRUE)
ew_girl_total <- sum(eng_wales5$Number[eng_wales5$Sex == "Girl"], na.rm = TRUE)

# Find North_Ireland total sex counts
ni_boy_total <- sum(nor_ire5$Number[nor_ire5$Sex == "Boy"], na.rm = TRUE)
ni_girl_total <- sum(nor_ire5$Number[nor_ire5$Sex == "Girl"], na.rm = TRUE)

# Find Scotland total sex counts
s_boy_total <- sum(scot5$Number[scot5$Sex == "Boy"], na.rm = TRUE)
s_girl_total <- sum(scot5$Number[scot5$Sex == "Girl"], na.rm = TRUE)

```

### Create a table displaying the data

```{r birth_summary_table}
# Build the basic table from the individual variables
birth_summary <- data.frame(
  Country = c("England & Wales", "England & Wales", "North Ireland", "North Ireland", "Scotland", "Scotland"),
  Sex     = c("Boy", "Girl", "Boy", "Girl", "Boy", "Girl"),
  Total   = c(ew_boy_total, ew_girl_total, ni_boy_total, ni_girl_total, s_boy_total, s_girl_total)
)

# Add the percentage values column to the basic table
birth_summary <- birth_summary %>%
  group_by(Country) %>%
  mutate(Percentage = (Total / sum(Total)) * 100)

# Use kable() to create the table in the GitHub markdown output
birth_summary %>%
  kable(
    col.names = c("Country", "Sex", "Total", "Percentage"),
    format.args = list(big.mark = ",")
 )
```

### Analysis Update

So, I was completely wrong about the possibility of there being more girls born in Scotland. In fact, according to the table above, there are more boys born than girls in each region across the 5 year time frame.

It looks like I'll need to do more investigation in the future to understand why the girls name hit the top ten in Scotland. 🥸🕵

## June 20, 2026

### Let's investigate girl names Olivia and Isla

I'm curious how many times the girls' names (Olivia and Isla) appear in each year.

#### Olivia

```{r countOlivia}
# Identify how many times Olivia appears
olivia <- filter(scot5, Name == "Olivia")
# Plot the Olivia data over time
olivia %>% ggplot(aes(x = Year, y = Rank)) + geom_col()
```

#### Isla

```{r countIsla}
# Identify how many times Isla appears
isla <- filter(scot5, Name == "Isla")
# Plot the Isla data over time
isla %>% ggplot(aes(x = Year, y = Rank)) + geom_col()
```


### Overview of findings

A quick view of the data frame shows the rank for both girls' names for each year, and they appear in at least the top 3 for each of those years, meaning they outrank the boys' names in each year except 2021, when Isla was 3rd. Although, Olivia was ranked 1st that year. 

Since we know that there were more boys born than girls in each of those years, it must mean that the boys' names were more dispersed than the girls' names. Two points of interest arise here:
### 1. why were those two names, Olivia and Isla, more popular for girl names, and
### 2. of the boys' names that made the top ranks, how disperse are their data?

## Chat with Gemini

### 1.
The popularity of the names Olivia and Isla may stem from pop culture with actresses, specifically Olivia Rodrigo,Olivia Colman, and Isla Fisher. It may also arise from the fictional character Isla Black Hitchens in the Harry Potter series.

[Source: Google. (2026). Gemini (June 20 version) [Large language model]. https://gemini.google.com/]

### 2.
Let's get an idea of the top boys names by rank. Since the boy name Noah ranked highest in our early Custom_Rank analysis, I want to find out which year his name ranked highest and explore that year further.

#### Investigate boy name Noah

```{r countNoah}
# Identify how many times Noah appears
noah <- filter(scot5, Name == "Noah")
# Plot the Noah data over time
noah %>% ggplot(aes(x = Year, y = Rank)) + geom_col()
```

### Even More Interesting!!!

The ranks show on a scale of hundreds up to 800, which meant that in the data Noah is ranked much further down the scale for 2021 and 2023. However, a quick view at the data frame shows the reason for this scale. In 2021 and 2023, there were a few girls named Noah! Since the rank is low (only 3 girls were named Noah in each of the years), I will eliminate the times Noah is ranked in the girls' names and rerun the plots.

#### Noah minus the girls

```{r noahNoGirls}
# Plot the Noah data over time
noah %>% filter(Sex == "Boy")  %>% ggplot(aes(x = Year, y = Rank)) + geom_col()
```

### Revelation

Interesting that Noah is ranked 2nd in 2020, 2021, and 2023 because Olivia is ranked 2nd in 2020 and 2023. Clearly the Rank variable in the original data set is stratified for girls' and boys' names. When I calculated the Custom_Rank, it included all names out of the total births for each year.

I need to rethink my strategy. 🤔
